Unsupervised Learning for Identifying Events in Active Target Experiments

### 2020-08-06 doc/Manuscript/article.tex
small fixes on references and changes to 5.1

@@

\section{Data Preparation}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.9\textwidth, height=9cm]{custom_work/examples_filtered.pdf}
\caption[Displaying filtered events in 2D and 3D]{Two- and three-dimensional representations of two events from the ${}^{46}$Ar experiment. Each row is one event in two projections, where the lightness of each point indicates higher charge values. These events have been filtered with a nearest neighbors algorithm and an algorithm based on the Hough transform \cite{Newman1972}.}
\end{figure}

@@

\begin{equation}\label{eq:ari}
\text{ARI} = \frac{\sum  \binom{w_{ij}}{2} - \left[\sum  \binom{e_{i}}{2} \sum  \binom{f_{j}}{2}  \right]/\binom{n}{2}}{\frac{1}{2}\left[\sum  \binom{e_{i}}{2} + \sum  \binom{f_{j}}{2}  \right]- \left[\sum  \binom{e_{i}}{2} \sum  \binom{f_{j}}{2}  \right]/\binom{n}{2}}.
\end{equation}
The quantities $f_i$ and $e_i$ are defined in table {tab:contingency}.
The important distinction to make between the clustering accuracy and the ARI is that the clustering accuracy is a simple comparison that does not account for chance assignments, while the ARI does. In effect, this means that the accuracy is a good heuristic for the performance, tempered by the ARI.

@@

With the best set of hyperparameters, each highest performing model is thereafter run $10$ times. The results are listed in table \ref{tab:clstr_mixae}.
We observe that, while the algorithm can achieve a very strong performance, the performance varies. In some cases the MIXAE model converges to a seemingly good configuration, based on its unsupervised training goals. However, when inspecting its clustering performance against labelled data, the seemingly good model does no better than a model based on random selection. 
This happens more frequently with the raw data, indicating an interaction with the noise levels present in the events.

% This table is not properly explained, what is init=10? and what is top 1? 
\begin{table}[H]
\centering 
\caption[MIXAE clustering performance]{MIXAE clustering performance on the ${}^{46} Ar$ experimental data. In contrast with the VGG-16 + $K$-means approach we observe  significant variations in performance. }\label{tab:clstr_mixae}
\input{./plots/mixae_clustering.tex}
\end{table}

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the $K$-means results or were significantly worse. Notably, the DBSCAN algorithm \cite{Ester96adensity-based,Bergstra2012} failed to provide any useful clustering results. We find this important as one of the significant drawbacks of $K$-means, and the deep clustering algorithm presented in section \ref{sec:mixae}, is that they depend  on pre-determining the number of clusters. This is not the case for DBSCAN.

@@

\section{Conclusions and Perspectives}\label{sec{conclusion}}
% Begin edit by DB (6/17/2020)
The purpose of this study has been to explore the application of unsupervised learning algorithms to event identification from an active target detector. The necessity to identify events from raw data prior to full processing is becoming a major issue in the data analysis of detectors with complex responses such as the AT-TPC.
As shown by both avenues explored in this work, it is clear that there is significant potential to eventually achieve event classification using fully automated unsupervised methods.

@@

%{\color{orange}[RS Another abrupt end to a paragraph. Could probably use some rounding out.]} 
\subsection*{Acknowledgements}
MHJ’s work is supported by U.S. National Science Foundation Grants No. PHY-1404159 and PHY-2013047.

>>>>

Within the context of machine learning methods applied to the analysis of nuclear physics experiments, 
the purpose of this work is thus to explore the application of unsupervised learning algorithms to event identification from an active target detector. The necessity to identify events from raw data prior to full processing is becoming a major issue in the data analysis of detectors with complex responses such as the AT-TPC.  After these introductory words, we review the experimental information in the next section. In section \ref{sec:datadescript} we describe the set up of the data while sections \ref{sec:me

@@

\section{Data Preparation}\label{sec:datadescript}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.9\textwidth, height=9cm]{custom_work/examples_filtered.pdf}
\caption[Displaying filtered events in 2D and 3D]{Two- and three-dimensional representations of two events from the ${}^{46}$Ar experiment. Each row is one event in two projections, where the lightness of each point indicates higher charge values. These events have been filtered with a nearest neighbors algorithm and an algorithm based on the Hough transform \cite{Newman1972}.}\label{fig:samples_filtered}
\end{figure}

@@

\begin{equation}\label{eq:ari}
\text{ARI} = \frac{\sum  \binom{w_{ij}}{2} - \left[\sum  \binom{e_{i}}{2} \sum  \binom{f_{j}}{2}  \right]/\binom{n}{2}}{\frac{1}{2}\left[\sum  \binom{e_{i}}{2} + \sum  \binom{f_{j}}{2}  \right]- \left[\sum  \binom{e_{i}}{2} \sum  \binom{f_{j}}{2}  \right]/\binom{n}{2}}.
\end{equation}
The quantities $f_i$ and $e_i$ are defined in table \ref{tab:contingency}.
The important distinction to make between the clustering accuracy and the ARI is that the clustering accuracy is a simple comparison that does not account for chance assignments, while the ARI does. In effect, this means that the accuracy is a good heuristic for the performance, tempered by the ARI.

@@

With the best set of hyperparameters, each highest performing model is thereafter run $10$ times\footnote{The MIXAE model is significantly more computationally expensive to train than a $K$-means model. Reulting in the skew in number of runs in the two cases.}. The results are listed in table \ref{tab:clstr_mixae}.
We observe that, while the algorithm can achieve a very strong performance, the performance varies. In some cases the MIXAE model converges to a seemingly good configuration, based on its unsupervised training goals. However, when inspecting its clustering performance against labelled data, the seemingly good model does no better than a model based on random selection. 
This happens more frequently with the raw data, indicating an interaction with the noise levels present in the events.

% This table is not properly explained, what is init=10? and what is top 1? 
\begin{table}[H]
\centering 
\caption[MIXAE clustering performance]{MIXAE clustering performance on the ${}^{46} Ar$ experimental data with $N=10$ runs of the algorithm. In contrast with the VGG-16 + $K$-means approach we observe  significant variations in performance.}\label{tab:clstr_mixae}
\input{./plots/mixae_clustering.tex}
\end{table}

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the Scikit-learn package of \citet{Pedregosa2011}. None of them provided any notable differences from the $K$-means results or were significantly worse. Notably, the DBSCAN algorithm \cite{Ester96adensity-based,Bergstra2012} failed to provide any useful clustering results. We find this important as one of the significant drawbacks of $K$-means, and the deep clustering algorithm presented in section \ref{sec:mixae}, is that they depend  on pre-determining the number of clusters. This is not the case for DBSCAN.

@@

\section{Conclusions and Perspectives}\label{sec:conclusion}
% Begin edit by DB (6/17/2020)
The purpose of this study has been to explore the application of unsupervised learning algorithms to event identification from an active target detector. The necessity to identify events from raw data prior to full processing is becoming a major issue in the data analysis of detectors with complex responses such as the AT-TPC.
As shown by both avenues explored in this work, it is clear that there is significant potential to eventually achieve event classification using fully automated unsupervised methods.

@@

%{\color{orange}[RS Another abrupt end to a paragraph. Could probably use some rounding out.]} 
\subsection*{Acknowledgements}
MHJ’s work is supported by U.S. National Science Foundation Grants No. PHY-1404159 and PHY-2013047. This project has also received support from the INTPART project of the Research Council of Norway (Grant No. 288125).
